Geographically aware estimates of remotely sensed water properties for Chesapeake Bay

### 2022-11-17 manuscript/manuscript.tex
include study site fig in supplement

Chesapeake Bay is located in the Mid-Atlantic region of the United States \hl{(at approximately 37.8\degree N 76.1\degree W)} and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River (at the Northern extent of the Bay) provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, \cite{xuClimateForcingSalinity2012}). \hl{The Bay is the largest estuary in the United States and has an area of approximately 11,000 $km^2$. As such, it is provides a wealth of natural economic benefits in the form of fisheries harvests (particularly oysters), tourism, and nature-based recreation} \cite{hoodChesapeakeBayProgram2021}. Despite its large size, the Bay is fairly shallow ($<$ 30m, \cite{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed \cite{xuClimateForcingSalinity2012}. For the present study, this means that surface water properties are likely to differ substantially from water properties at deeper depths. It also means that the Bay receives enough freshwater inputs to produce a distinct upper-estuary to lower-estuary gradient in water properties.

@@

Our tuned RF models for salinity, temperature, and turbidity had a root mean square error (RMSE) of 2.34 ($R^2$=0.91), 1.76 ($R^2$=0.96), and 1.82 ($R^2$=0.11) respectively (Table 1, Figure 2). Out-of-sample validation performance was good for salinity and temperature models but poor for the turbidity model (Figure 2, Figure S1). Despite excellent overall performance for the temperature and salinity models, there was some indication of poor performance at the extremes. The models were unable to recover very low temperature values ($<$ 0 C) or very high temperature and salinity values ($>$ 30 C, $>$ 23 salinity, Figure 2). One explanation for these areas of poor performance is that such extremes were rare in the observational dataset (Table S2). The salinity model however did faithfully capture areas of low salinity (Figure 2, S3). These low salinity values can be regarded as low uncertainty (high confidence) estimates given the extent to which the upper tributaries are well represented in the observational dataset (Figure S2).

@@

Our comprehensive dataset includes greater spatial coverage than prior efforts that were  largely trained on limited (and relatively high-salinity) buoy or in-situ cruise data located in the mainstem of the Bay \cite{vogelAssessingSatelliteSea2016, geigerSatellitederivedCoastalOcean2013, ondrusekDevelopmentNewOptical2012}. This coverage allows for a more comprehensive assessment of the spatial uncertainty of our model. Predicted areas of low salinity coming from our model likely have particularly low associated uncertainties because the upper tributaries were well-represented in our observational dataset (Figure S2). Because CBOFS salinity data was fresher than our RF estimates in the upper tributaries, this suggests that CBOFS data for these areas are likely true underestimates and not a deficiency of the RF model predictions.

In contrast to the upper tributaries, an area of maximum uncertainty is likely the mainstem and eastern Bay across from the mouth of the Potomac (Tangier Sound). This appears to be the collision point of the marine plume emanating from the Bay mouth with the discharge of the Potomac River. The hydrodynamics here seem to be complex given that the CBOFS has difficulty resolving spatial patterning in this area \cite{lanerolle2011second}. This was an area where we found limited observational data (Figure S2). Improvements to data-driven modeling in this area may be possible with the inclusion of more observational data.

>>>>

Chesapeake Bay is located in the Mid-Atlantic region of the United States \hl{(at approximately 37.8\degree N 76.1\degree W, see Figure S1)} and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River (at the Northern extent of the Bay) provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, \cite{xuClimateForcingSalinity2012}). \hl{The Bay is the largest estuary in the United States and has an area of approximately 11,000 $km^2$. As such, it is provides a wealth of natural economic benefits in the form of fisheries harvests (particularly oysters), tourism, and nature-based recreation} \cite{hoodChesapeakeBayProgram2021}. Despite its large size, the Bay is fairly shallow ($<$ 30m, \cite{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed \cite{xuClimateForcingSalinity2012}. For the present study, this means that surface water properties are likely to differ substantially from water properties at deeper depths. It also means that the Bay receives enough freshwater inputs to produce a distinct upper-estuary to lower-estuary gradient in water properties.

@@

Our tuned RF models for salinity, temperature, and turbidity had a root mean square error (RMSE) of 2.34 ($R^2$=0.91), 1.76 ($R^2$=0.96), and 1.82 ($R^2$=0.11) respectively (Table 1, Figure 2). Out-of-sample validation performance was good for salinity and temperature models but poor for the turbidity model (Figure 2, Figure S2). Despite excellent overall performance for the temperature and salinity models, there was some indication of poor performance at the extremes. The models were unable to recover very low temperature values ($<$ 0 C) or very high temperature and salinity values ($>$ 30 C, $>$ 23 salinity, Figure 2). One explanation for these areas of poor performance is that such extremes were rare in the observational dataset (Table S2). The salinity model however did faithfully capture areas of low salinity (Figure 2, S4). These low salinity values can be regarded as low uncertainty (high confidence) estimates given the extent to which the upper tributaries are well represented in the observational dataset (Figure S3).

@@

Our comprehensive dataset includes greater spatial coverage than prior efforts that were  largely trained on limited (and relatively high-salinity) buoy or in-situ cruise data located in the mainstem of the Bay \cite{vogelAssessingSatelliteSea2016, geigerSatellitederivedCoastalOcean2013, ondrusekDevelopmentNewOptical2012}. This coverage allows for a more comprehensive assessment of the spatial uncertainty of our model. Predicted areas of low salinity coming from our model likely have particularly low associated uncertainties because the upper tributaries were well-represented in our observational dataset (Figure S3). Because CBOFS salinity data was fresher than our RF estimates in the upper tributaries, this suggests that CBOFS data for these areas are likely true underestimates and not a deficiency of the RF model predictions.

In contrast to the upper tributaries, an area of maximum uncertainty is likely the mainstem and eastern Bay across from the mouth of the Potomac (Tangier Sound). This appears to be the collision point of the marine plume emanating from the Bay mouth with the discharge of the Potomac River. The hydrodynamics here seem to be complex given that the CBOFS has difficulty resolving spatial patterning in this area \cite{lanerolle2011second}. This was an area where we found limited observational data (Figure S3). Improvements to data-driven modeling in this area may be possible with the inclusion of more observational data.

### 2022-11-17 manuscript/supplement.tex
move validation plot from supp to ms
formally include the obs_distribution table
text edits
prep pdf ms
include study site fig in supplement

@@

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=0.7\textwidth,keepaspectratio]{figures/_validation}
          \caption{RF predicted salinity versus test set (observed) salinity. Hatching indicates the area containing 80\% of the data points.}
    \end{center}    
\end{figure}

@@

>>>>

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=0.85\textwidth,keepaspectratio]{figures/_study_site}
          \caption{Map of Cheseapeake Bay study site.}
    \end{center}    
\end{figure}

@@

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=0.85\textwidth,keepaspectratio]{figures/obs_distribution_table}
    \end{center}    
\end{figure}

@@

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=1\textwidth,keepaspectratio]{figures/_seasonality_turbidity}
          \caption{Seasonal Random Forest prediction results for turbidity.}
    \end{center}    
\end{figure}

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=1\textwidth,keepaspectratio]{figures/_freqcount_hex}
          \caption{Locations of observational data.}
    \end{center}    
\end{figure}

\begin{figure}[h!]
    \begin{center}
          \includegraphics[width=1\textwidth,keepaspectratio]{figures/_validation_map}
          \caption{Map of model validation errors.}
    \end{center}    
\end{figure}
